---
title: "Firstpass_Biomass"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: inline
---

# Estimating Total Aboveground Biomass

```{r include=FALSE}
# turns off annoying "unknown column" warning
# !diagnostics off 

library(ggplot2)
library(cowplot); theme_set(theme_cowplot())
library(tidyverse)
library(here)
options(scipen = 999)
```


```{r}
dbh <- read.csv(here("data/dbh.csv"))

dbh %>%
  group_by(SPP) %>%
  count()
```

I'm skipping balsam poplar for now (abbreviated as POBA), since there's only 3 of them in the whole dataset. Plus, Yarie et al. 2007 doesn't work for it, so I'll have to track down equations based on smaller individuals. 

I've also skipped the shrubs for now - Berner et al. 2015 has equations for arctic birch (ARCTO in the datafile), willow and alder (ALCR), but they depend on basal area, not DBH. Going to keep looking.

## Birch
I've been compiling each species individually, starting with birch. Yarie et al. provides both equations for total aboveground biomass on its own, as well as equations for each component. I give both a shot below, choosing only the equations with the higher r2 value. I also include the standard error values for each equation, and carry them through to a final estimate. 

```{r}
# selecting just for birch
birch <- dbh %>%
  filter(SPP == "BENE")
# summary(birch$DBH) # min 0.1, max 23.5
hist(birch$DBH) # big stuff (dbh > 18) is all unburned sites (dalton 58_0 and 10_0, steese 6_0), all dead
```

```{r}
##########################
# testing total aboveground biomass equations

# first equation in table
EQ_tab_bene <- function(x) -16809 + (2656.482*x) + (284.416*(x^2))
# n = 19, r2 = 0.98
# SE
  # intercept = 10934
  # dbh = 1462.65
  # dbh2 = 42.72
    tab_beneSE <- sqrt((10934^2) + (1462.65^2) + (42.72^2))

birch$EQ_tab <- EQ_tab_bene(birch$DBH)

#############################
# testing sum of component equations

# FOLIAGE
fol_bene  <- function(x) 145.892*x
# n = 19, r2 = 0.877
# SE
  # dbh = 12.89
    fol_beneSE <- sqrt(12.89^2)

birch$fol <- fol_bene(birch$DBH)

# LIVE BRANCHES
branch_bene <- function(x) -472.032*x + 81.536*(x^2)
# n = 19, r2 = 0.949
# SE
  # dbh = 202.38
  # dbh2 = 9.25
    branch_beneSE <- sqrt((202.38)^2  + (9.25)^2)

birch$branch <- branch_bene(birch$DBH)

# STEMWOOD
stemw_bene <- function(x) 1783.003*x + 177.304*(x^2)
# n = 11, r2 = 0.978
  # SE
    # dbh = 860.5
    # dbh2 = 35.99
      stemw_beneSE <- sqrt(860.5^2 + 35.99^2)

birch$stem <- stemw_bene(birch$DBH)

# BARK
bark_bene <- function(x) 49.01*(x^2)
# n = 10, r2 = 0.992
# SE
  # dbh2 = 1.44
    bark_beneSE <- sqrt(1.44^2)

birch$bark <- bark_bene(birch$DBH)

# summing up components
birch$EQ_sumR <- birch$fol + birch$branch + birch$stem + birch$bark

sumR_beneSE <- sqrt(fol_beneSE^2 + branch_beneSE^2 + stemw_beneSE^2 + bark_beneSE^2)

# reformating dataset
birch <- birch %>%
  pivot_longer(
    cols = starts_with("EQ"),
    names_to = "EQ",
    values_to = "Biomass")

# clean up variable name
birch$EQ[birch$EQ == "EQ_tab"] <- "tab"
birch$EQ[birch$EQ == "EQ_sumR"] <- "sumR"

# adding in standard errors
birch$SE <- NA
birch$SE[birch$EQ == "tab"] <- tab_beneSE
birch$SE[birch$EQ == "sumR"] <- sumR_beneSE

# taking out component columns to clean up dataset
birch <- subset(birch, select = -c(fol, bark, stem, branch))
rm(bark_beneSE, branch_beneSE, fol_beneSE, stemw_beneSE )
```

Three notes about the graph below: 
1. The units on the plot below aren't the final units: trees in some plots were counted across smaller subsets because of density, so the dataset contains a column called "Expansion factor" that I'll use a little later to scale everything up. 

2. Calculating total aboveground biomass of birch using the sum of each component actually provides a negative estimate of biomass at smaller DBHs (hence the line that cuts off the graph at y = 0). I may eventually look for an equation based more firmly on smaller individuals, but I'll stick to using the total aboveground biomass equation for now, since those values aren't negative. 

3. I don't think the uncertainty estimates here are calculated right (especially the sum of components), but we're putting a pin in that for now. 

```{r}
ggplot(birch, aes(x = DBH)) +
  geom_point(aes(y = Biomass, col = EQ)) +
  geom_ribbon(aes(ymax = Biomass + SE, ymin = Biomass - SE, fill = EQ), alpha = .3) +
  geom_vline(xintercept = c(0.1, 23.5), linetype = "dashed") +
  geom_vline(xintercept = c(3.7, 32.7), linetype = "dashed", col = "blue")  +
  scale_color_manual(values = c("red", "blue"),
                     labels = c("Sum of Components", "Tot. Above. Biomass"),
                     name = "Equation") +
  scale_fill_manual(values = c("red", "blue"),
                    labels = c("Sum of Components", "Tot. Above. Biomass"),
                    name = "Equation") +
  labs(x = "DBH (cm)", y = "Dry Biomass (g)", title = "Alaskan Birch (Total Above. Biomass)") +
  geom_text(mapping = aes(x = 1.5, y = 185000, label = "Observed Range"),
            angle = 90, size = 3, check_overlap = TRUE) +
  geom_text(mapping = aes(x = 5, y = 185000, label = "Equation Range"),
            angle = 90, size = 3, color = "Blue", check_overlap = TRUE) +
  ylim(c(0, 225000))
```

```{r}
# selecting for just total aboveground biomass
birch <- birch %>%
  filter(EQ == "tab")
  # getting rid of extra column
    birch <- subset(birch, select = - EQ)
    rm(sumR_beneSE, tab_beneSE)
```


## Black Spruce

I tracked down the largest trees in the histogram below - there's 2 trees with a DBH over 20 cm. One's a large live tree in an unburned site (23.6 cm, dalton, 58_0), the other's a dead tree in a once-burned site (26 cm, steese, 18_1). 

```{r}
pime <- dbh %>%
  filter(SPP == "PIME")
summary(pime$DBH) # min 0.1, max 25
hist(pime$DBH)
```

I'm using the Yarie equations here too, with the same approach as above - picking the largest r2 values and testing out both the total aboveground equation and the equation based on the sum of components. 

```{r}
##########################
# testing total aboveground biomass equations

tab_pime <- function(x)  358.352*x + 158.166*(x^2)
# n = 78, r2 = 0.905
# SE
# dbh = 240.81
# dbh2 = 26.75
tab_pimeSE <- sqrt((240.81)^2 + (26.75)^2)

pime$EQ_tab <- tab_pime(pime$DBH)

########################
# calculating each component

# FOLIAGE
fol_pime  <- function(x) 161.583*x
# n = 56, r2 = 0.680
# SE
# dbh = 14.95
fol_pimeSE <- sqrt(14.95^2)

pime$fol <- fol_pime(pime$DBH)

# LIVE BRANCHES
branch_pime <- function(x) 98.535*x + 7.693*(x^2)
# n = 56, r2 = 0.685
# SE
# dbh = 43.83
# dbh2 = 4.98
branch_pimeSE <- sqrt(43.83^2 + 4.98^2)

pime$branch <- branch_pime(pime$DBH)

# STEMWOOD
stemw_pime <- function(x) -267.688*x + 159.822*(x^2)
# n = 9, r2 = 0.998
# SE
# dbh = 128.39
# dbh2 = 11.34
stemw_pimeSE <- sqrt(128.39^2 + 11.34^2)

pime$stem <- stemw_pime(pime$DBH)

# BARK
bark_pime <- function(x) 26.134*(x^2)
# n = 9, r2 = 0.998
# SE
# dbh2 = 0.4
bark_pimeSE <- sqrt(0.4^2)

pime$bark <- bark_pime(pime$DBH)

# summing up components
pime$EQ_sumR <- pime$fol + pime$branch + pime$stem + pime$bark

sumR_pimeSE <- sqrt(fol_pimeSE^2 + branch_pimeSE^2 + stemw_pimeSE^2 +  bark_pimeSE^2)

# pivoting out
pime <- pime %>%
  pivot_longer(
    cols = starts_with("EQ"),
    names_to = "EQ",
    values_to = "Biomass")

# clean up variable name
pime$EQ[pime$EQ == "EQ_tab"] <- "tab"
pime$EQ[pime$EQ == "EQ_sumR"] <- "sumR"

# adding in standard errors
pime$SE <- NA
pime$SE[pime$EQ == "tab"] <- tab_pimeSE
pime$SE[pime$EQ == "sumR"] <- sumR_pimeSE

pime <- subset(pime, select = -c(fol, bark, stem, branch))
```


Thankfully, the Yarie equations here don't produce any negative values. I think the same issue with uncertainty is happening here too, but I'm ignoring that for now. The two equations definitely start to diverge above DBHs of 15 cm or so, so I may eventually try to find something that used individuals across a wider range. For now, picking the total aboveground equation for simplicity (same as birch), and in order to underestimate, not overestimate. 

```{r}
ggplot(pime, aes(x = DBH)) +
  geom_point(aes(y = Biomass, col = EQ)) +
  geom_ribbon(aes(ymax = Biomass + SE, ymin = Biomass - SE, fill = EQ), alpha = .3) +
  geom_vline(xintercept = c(0.1, 26), linetype = "dashed") +
  geom_vline(xintercept = c(0, 12.8), linetype = "dashed", col = "blue")  +
  scale_color_manual(values = c("red", "blue"),
                     labels = c("Sum of Components", "Tot. Above. Biomass"),
                     name = "Equation") +
  scale_fill_manual(values = c("red", "blue"),
                    labels = c("Sum of Components", "Tot. Above. Biomass"),
                    name = "Equation") +
  labs(x = "DBH (cm)", y = "Dry Biomass (g)", title = "Black Spruce (Total Above. Biomass)") +
  geom_text(mapping = aes(x = 1.5, y = 90000, label = "Observed Range"),
            angle = 90, size = 3, check_overlap = TRUE) +
  geom_text(mapping = aes(x = 14, y = 90000, label = "Equation Range"),
            angle = 90, size = 3, color = "Blue", check_overlap = TRUE) +
  ylim(c(0, 110000))
```
```{r}
# going with total aboveground for consistency
pime <- pime %>%
  filter(EQ == "tab")
pime <- subset(pime, select = - c(EQ))
rm(bark_pimeSE, branch_pimeSE, fol_pimeSE, stemw_pimeSE, sumR_pimeSE, tab_pimeSE)
```

## Aspen

The Yarie equations almost entirely produced negative estimates for aspen, since ours were so small and theirs were so large (max of 44 cm, which is massive compared to our sites). I ended up going with the Ung et al. 2008 equation. They didn't create a total aboveground equations, only the components, so I'm only using that one. 

```{r}
  potr <- dbh %>%
    filter(SPP == "POTR")
  # summary(potr$DBH) # min 0.1, max 6.5
  hist(potr$DBH) # everything smaller than 7 cm
```

```{r}
########################
  # calculating each component

  # FOLIAGE
  fol_potr <- function(x) 0.0235*x^(1.6656)
  # n = ,r2 = 0.42
  # SE
  # intercept = 0.0032
  # dbh^ = 0.0440
  fol_potrSE <- sqrt((0.0032^2) + (0.0440^2))

  potr$fol <- fol_potr(potr$DBH)

  # LIVE BRANCHES
  branch_potr <- function(x)  0.0082*x^(2.5139)
  # n = , r2 = 0.76
  # SE
  # intercept = 0.0008
  # dbh^ = 0.0327
  branch_potrSE <- sqrt((0.0008^2) + (0.0327^2))

  potr$branch <- branch_potr(potr$DBH)

  # WOOD
  wood_potr <- function(x) 0.0608*x^(2.4735)
  # n = , r2 =0.95
  # SE
  # intercept = 0.0029
  # dbh^ = 0.0153
  wood_potrSE <- sqrt((0.0029^2) + (0.0153^2))

  potr$wood <- wood_potr(potr$DBH)

  # BARK
  bark_potr <- function(x) 0.0159*x^(2.4123)
  # n = , r2 = 0.91
  # SE
  # intercept = 0.0006
  # dbh^ = 0.0131
  bark_potrSE <- sqrt((0.0006^2) + (0.0131^2))

  potr$bark <- bark_potr(potr$DBH)

  # summing up components
  potr$sum <- potr$fol + potr$branch + potr$wood + potr$bark

  # standard error
  potr$SE <- sqrt(fol_potrSE^2 + branch_potrSE^2 + wood_potrSE^2 + bark_potrSE^2)
```

There's only one equation, so the graph below is a little overkill, but at least it shows there's no negative values and the range of DBH is relatively well-represented. Still think uncertainty is wrong here, still ignoring for now. 

```{r}
ggplot(potr, aes(x = DBH, y = sum)) +
    geom_point() +
    geom_ribbon(aes(ymax = sum + SE, ymin = sum - SE), alpha = .3) +
    labs(x = "DBH (cm)", y = "Dry Biomass (g)", title = "Trembling Aspen (Total Above. Biomass)") +
    geom_vline(xintercept = c(0.1, 6.5), linetype = "dashed") +
    geom_vline(xintercept = c(0.7, 47.2), linetype = "dashed", col = "blue")  +
    xlim(0, 8) +
    geom_text(mapping = aes(x = 0.35, y = 8, label = "Observed Range"),
              angle = 90, size = 3) +
    geom_text(mapping = aes(x = 1, y = 8, label = "Equation Range"),
```
```{r}
  # renaming columns
  potr <- potr %>%
    rename("Biomass" = "sum")
  # removing each component
  potr <- subset(potr, select = -c(fol, branch, wood, bark))
  rm(bark_potrSE, branch_potrSE, fol_potrSE, wood_potrSE)
```

# Combining species values

```{r}
tab <- rbind(pime, birch, potr)

  summary(tab$Biomass)  # so we still ended up with negative biomass estimates, not great

  # ignoring negative values for now
  tab <- tab %>%
    filter(Biomass > 0)

  summary(tab$Biomass)

  # scaling up by expansion factors
  tab$biomass_ha <- tab$EXP_FACT * tab$Biomass # here, Biomass = per plot 
  tab$biomass_m2 <- tab$biomass_ha / 10000

  # grouping at plot level
  tab_plot <- tab %>%
    group_by(SITE, TREAT, PLOT, SPP) %>%
    summarise(Biomass = sum(biomass_m2)) %>%
    ungroup() %>%
    complete(nesting(SITE, TREAT, PLOT), SPP, fill = list(Biomass = 0))

    tab_plot$Biomass <- round(tab_plot$Biomass, digits = 4)
```


```{r}
ggplot(tab_plot, aes(x = as.factor(TREAT), y = Biomass)) + geom_boxplot() +
  labs(x = "Number of Fires", y = "Biomass (g/m2)", title = "Total Aboveground Biomass per plot")

```

```{r}
ggplot(tab_plot, aes(x = as.factor(TREAT), y = Biomass, fill = SPP)) + geom_boxplot() +
  labs(x = "Number of Fires", y = "Biomass (g/m2)", title = "Biomass per species")
```
Assuming those values look okay, the next steps might be something like:
- carry canopy values through to total plot estimates so I can plot live vs dead. 
- fix the issue with uncertainty - I'm sure those numbers should be larger. 
- track down shrub equations that work with DBH. 

