---
title: "Biomass Model"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: inline
---

# Biomass model - 12/1
```{r include=FALSE, packages}
library(tidyverse)
library(R2jags)
library(UsingR)
library(GGally) # correlation plots
library(here)
library(car)
```

Loading in data - all variables are summed at the plot level, since % deciduous can only exist at the plot level (losing some detail in organic layer depth, density and biomass, but).

I'm also excluding unburned sites. 

```{r set up, echo=TRUE}
data <- read.csv(here("data/output/biomass_variables.csv"))

data <- data %>%
  dplyr::filter(TREAT != 0)

head(data)
```

Checking out some correlation plots beforehand. 
```{r Correlation}
ggally::ggpairs(data, columns = c(2,4:7),
        ggplot2::aes(fill = SITE, color = SITE),
        columnLabels = c("Fires", "Biomass (Kg/Ha)", "Density (Tree/Ha)",
                       "% Deciduous", "Org. Layer depth (cm)")) +
  scale_fill_manual(values = c("#7bccc4", "#fe9929"),
                    labels = c("Upland", "Lowland")) +
  scale_color_manual(values = c("#7bccc4", "#fe9929"),
                     labels = c("Upland", "Lowland"))
```

# Checking distributions
```{r density plots}
DensityPlot(data$Biomass_Kg_ha)
DensityPlot(data$Count_HA)
DensityPlot(data$per_d)
DensityPlot(data$AV_DEPTH)
```

```{r}
qqPlot(data$Biomass_Kg_ha)
qqPlot(data$Count_HA)
qqPlot(data$per_d)
qqPlot(data$AV_DEPTH)
```

```{r}
qqPlot(log(data$Biomass_Kg_ha))
qqPlot(log(data$Count_HA))
```

# Density model

I'm starting with a 'basic' model of density, assuming normality (which, per the qqplots, isn't safe). 

## Linear Model
Here's what the GLM looks like: 
```{r density Linear Model, echo=TRUE}
glm_density <- glm(Biomass_Kg_ha ~ Count_HA + SITE, data = data)

  summary(glm_density)
```

```{r density linear model plot}
  plot(Biomass_Kg_ha ~ Count_HA, data = data)
  abline(glm_density, col = "red")
```

### Setting up JAGS
```{r density data linear uniformative}
 # defining data 
  y_biomass <- as.vector(data$Biomass_Kg_ha)
  x_density <- as.vector(data$Count_HA)
  x_site <- as.vector(data$SITE)
  x_site <- as.factor(x_site) # won't work if not a factor
```

The model is based on biomass = density + site, with site acting as a binary variable.

```{r density model jags linear uniformative}
 dens_model1_jags <- function(){
  # Likelihood:
  for (i in 1:N){
    y_biomass[i] ~ dnorm(beta0 + beta1*x_density[i] + beta2*x_site[i], invsigma2)
  }
  # Priors:
  beta0 ~ dnorm(mu0, g0) # intercept
  beta1 ~ dnorm(mu1, g1) # change in biomass per increase in density
  beta2 ~ dnorm(mu2, g2) # strength of effect of site
  invsigma2 ~ dgamma(a, b) # all other processes 
}
```

```{r density initializing linear uniformative}
  dens_jagsdata_1 <- with(data, # creates a list of three elements
                    list(y_biomass = y_biomass,
                         x_density = x_density,
                         x_site = x_site,
                         N = length(data$Biomass_Kg_ha),
                         mu0 = 0, g0 = 0.0025, # using uninformative priors
                         mu1 = 0, g1 = 0.0025, # 
                         mu2 = 0, g2 = 0.0025,
                         a = 0.001, b = 0.001))

# specifies initial parameter values
init_values <- function(){
  list(beta0 = rnorm(1), beta1 = rnorm(1), beta2 = rnorm(1), invsigma2 = runif(1))
}

params <- c("beta0", "beta1", "beta2", "invsigma2") # chooses the parameters for which to report posteriors
```

### Run Model
```{r density run model linear uniformative, echo=TRUE}
dens_fit_lm1 <- jags(data = dens_jagsdata_1, # specifies data
                inits = init_values, # specifies initial values
                parameters.to.save = params,
                model.file = dens_model1_jags,
                n.chains = 3, # number of chains
                n.iter = 12000, # number of iterations
                n.burnin = 6000, # discards the first 2k values
                n.thin = 10) # only keeps every 10th iteration
```

### Diagnostics
```{r density diagnostics linear uniformative}
dens_fit_lm1 # model output

traceplot(dens_fit_lm1, mfrow = c(2, 2), ask = F) # traceplots of all three chains

  plot(dens_fit_lm1)

  dens_model1_mcmc <- as.mcmc(dens_fit_lm1) # need to convert to class 'mcmc' first
  plot(dens_model1_mcmc) # complete posterior distributions
  # traceplots + densities
```

## 2. Logged Linear model
### 2.1 Setting up JAGS
```{r density log linear uniformative set up}
y_biomass_log <- log(y_biomass)
x_density_log <- log(x_density)
```

```{r density model log linear uniformative}
dens_model2_jags <- function(){
  # Likelihood:
  for (i in 1:N){
    y_biomass_log[i] ~ dnorm(beta0 + beta1*x_density_log[i] + beta2*x_site[i], invsigma2)
  }
  # Priors:
  beta0 ~ dnorm(mu0, g0) # intercept
  beta1 ~ dnorm(mu1, g1) #
  beta2 ~ dnorm(mu2, g2) #
  invsigma2 ~ dgamma(a, b)
}
```

```{r density initialization log linear uniformative}
# specifies initial parameter values
init_values <- function(){
  list(beta0 = rnorm(1), beta1 = rnorm(1), beta2 = rnorm(1), invsigma2 = runif(1))
}

params <- c("beta0", "beta1", "beta2", "invsigma2") # chooses the parameters for which to report posteriors

dens_jagsdata_2 <- with(data, # creates a list of three elements
                    list(y_biomass_log = y_biomass_log,
                         x_density_log = x_density_log,
                         x_site = x_site,
                         N = length(data$Biomass_Kg_ha),
                         mu0 = 0, g0 = 0.0025,
                         mu1 = 0, g1 = 0.0025,
                         mu2 = 0, g2 = 0.0025,
                         a = 0.001, b = 0.001))
```

### 2.2 Run Model
```{r run density model log linear uniformative}
dens_fit_model2 <- jags(data = dens_jagsdata_2, # specifies data
                inits = init_values, # specifies initial values
                parameters.to.save = params,
                model.file = dens_model2_jags,
                n.chains = 3, # number of chains
                n.iter = 12000, # number of iterations
                n.burnin = 6000, # discards the first 2k values
                n.thin = 10) # only keeps every 10th iteration
```


### 2.3 Diagnostics
```{r density diagnostics log linear uniformative}
dens_fit_model2 # model output

traceplot(dens_fit_model2, mfrow = c(2, 2), ask = F) # traceplots of all three chains

  plot(dens_fit_model2)

  lm1_mcmc <- as.mcmc(dens_fit_model2) # need to convert to class 'mcmc' first
  plot(dens_fit_model2) # complete posterior distributions
  # traceplots + densities
```

# Deciduous Model
## A. Linear Model
```{r deciduous linear Model, echo=TRUE}
decid_glm <- glm(Biomass_Kg_ha ~ per_d + SITE, data = data)

  summary(decid_glm)
  plot(Biomass_Kg_ha ~ Count_HA, data = data)
  abline(decid_glm, col = "red")
```

### A.1 Setting up JAGS
```{r deciduous data linear uniformative}
  x_decid <- as.vector(data$per_dec)
```

```{r decid model linear uniformative}
 decid_model1_jags <- function(){
  # Likelihood:
  for (i in 1:N){
    y_biomass[i] ~ dnorm(beta0 + beta1*x_decid[i] + beta2*x_site[i], invsigma2)
  }
  # Priors:
  beta0 ~ dnorm(mu0, g0) # intercept
  beta1 ~ dnorm(mu1, g1) #
  beta2 ~ dnorm(mu2, g2) #
  invsigma2 ~ dgamma(a, b)
}
```

```{r decid initializing linear uniformative}
  decid_jagsdata_1 <- with(data, # creates a list of three elements
                    list(y_biomass = y_biomass,
                         x_decid = x_decid,
                         x_site = x_site,
                         N = length(data$Biomass_Kg_ha),
                         mu0 = 0, g0 = 0.0025,
                         mu1 = 0, g1 = 0.0025,
                         mu2 = 0, g2 = 0.0025,
                         a = 0.001, b = 0.001))

# specifies initial parameter values
init_values <- function(){
  list(beta0 = rnorm(1), beta1 = rnorm(1), beta2 = rnorm(1), invsigma2 = runif(1))
}

params <- c("beta0", "beta1", "beta2", "invsigma2") # chooses the parameters for which to report posteriors
```

### A. 2 Run Model
```{r decid run model linear uniformative}
decid_fit_1 <- jags(data = decid_jagsdata_1, # specifies data
                inits = init_values, # specifies initial values
                parameters.to.save = params,
                model.file =  decid_model1_jags,
                n.chains = 3, # number of chains
                n.iter = 12000, # number of iterations
                n.burnin = 6000, # discards the first 2k values
                n.thin = 10) # only keeps every 10th iteration
```

### A. 3 Diagnostics
```{r decid diagnostics linear uninformative}
decid_fit_1 # model output

traceplot(decid_fit_1, mfrow = c(2, 2), ask = F) # traceplots of all three chains

  plot(decid_fit_1)

  decid1_mcmc <- as.mcmc(decid_fit_1) # need to convert to class 'mcmc' first
  plot(decid1_mcmc) # complete posterior distributions
  # traceplots + densities
```

## B. Logged Linear model
### B. 1 Setting up JAGS
```{r}
y <- as.vector(data$Biomass_Kg_ha)
y <- log(y)
x_density <- as.vector(data$Count_HA)
x_density <- log(x_density)
x_site <- as.vector(data$SITE)
x_site <- as.factor(x_site)
```

```{r Run Model}
lm1_jags <- function(){
  # Likelihood:
  for (i in 1:N){
    y[i] ~ dnorm(beta0 + beta1*x_density[i] + beta2*x_site[i], invsigma2)
  }
  # Priors:
  beta0 ~ dnorm(mu0, g0) # intercept
  beta1 ~ dnorm(mu1, g1) #
  beta2 ~ dnorm(mu2, g2) #
  invsigma2 ~ dgamma(a, b)
}
```

```{r initialization}
# specifies initial parameter values
init_values <- function(){
  list(beta0 = rnorm(1), beta1 = rnorm(1), beta2 = rnorm(1), invsigma2 = runif(1))
}

params <- c("beta0", "beta1", "beta2", "invsigma2") # chooses the parameters for which to report posteriors

jagsdata_s1 <- with(data, # creates a list of three elements
                    list("y" = y,
                         "x_density" = x_density,
                         "x_site" = x_site,
                         "N" = length(data$Biomass_Kg_ha),
                         "mu0" = 0, "g0" = 0.0025,
                         "mu1" = 0, "g1" = 0.0025,
                         "mu2" = 0, "g2" = 0.0025,
                         "a" = 0.001, "b" = 0.001))
```

### B. 2 Run Model
```{r}
fit_lm1 <- jags(data = jagsdata_s1, # specifies data
                inits = init_values, # specifies initial values
                parameters.to.save = params,
                model.file = lm1_jags,
                n.chains = 3, # number of chains
                n.iter = 12000, # number of iterations
                n.burnin = 6000, # discards the first 2k values
                n.thin = 10) # only keeps every 10th iteration
```


### B. 3 Diagnostics
```{r}
fit_lm1 # model output

traceplot(fit_lm1, mfrow = c(2, 2), ask = F) # traceplots of all three chains

  plot(fit_lm1)

  lm1_mcmc <- as.mcmc(fit_lm1) # need to convert to class 'mcmc' first
  plot(lm1_mcmc) # complete posterior distributions
  # traceplots + densities
```

```{r}

```


```{r}

```

```{r}

```

